Data Specialization for a Staged Lambda Calculus

### 2015-02-12 paper/semantics.tex
worked on 3.0-3.1

The input to our splitting algorithm is given in \lang, a typed, modal
lambda calculus with products, sums, and isorecursive types. This language is
adapted from \cite{davies96}, and is very similar to the staged languages
considered in the partial evaluation and metaprogramming literature,
though it differs in three key ways:

@@

\item We restrict our attention to a two-stage language to simplify our
presentation of the splitting algorithm, though we expect that the techniques
presented here extend to more stages.

\item Although our language describes computations with two {\em stages}
(\bbone\ and \bbtwo), its type system makes use of three {\em worlds}
(\bbonem, \bbonep, \bbtwo).  For reasons discussed later, we require
that stage \bbone\ computations be classified as being either pure
(occuring at \bbonep) or mixed (occuring at \bbonem). 
Only mixed code may transition to stage/world \bbtwo.

@@

%Apart from the non-duplication, the dynamic semantics of \lang\ is essentially
%identical to Davies'.

@@

To ensure that the staging features are only applied in a valid way, 
we index our judgments not only by a type, but also a stage $w$.
The typing judgment $\typeswor e A$ says that $e$ has type $A$ at stage $w$, in
the context $\Gamma$, and is defined in \cref{fig:statics}.

\crem{Notice that certain types are impossible.}

Except for the staging features $\next$, $\prev$, and $\pause$, the ordinary
language features (i.e., the introductory and eliminatory forms for product,
sum, and function types) preserve the stage of their subterms, and work at both
stages \bbone\ and \bbtwo. Thus, if we were to remove the staging features,
\lang\ would simply consist of two non-interacting copies of a standard lambda
calculus. Variables in the context are annotated with the stage at which they
were introduced and can only be used at the same stage.

Note that there is no way to eliminate a term of type $\fut A$ into a stage
\bbone\ term---the only elimination form is $\prev$, which results in a stage
\bbtwo\ term. This is why well-typed terms cannot have any information flow
from stage \bbtwo\ to stage \bbone, and what makes splitting possible.

>>>>

We express staged programs as terms in \lang, a typed, modal lambda calculus
with products, sums, and isorecursive types. This language is adapted from
\cite{davies96}, and is very similar to other staged languages considered in the
partial evaluation and metaprogramming literature, though it differs in three
key ways:

@@

\item We restrict \lang\ to exactly two stages, to simplify our
presentation of our stage-splitting algorithm. We expect that the techniques
presented here extend to more stages.

\item Although \lang\ describes two-stage computations, its modal type system
has three distinct \emph{worlds} \bbonem, \bbonep, and \bbtwo. Essentially, this
further distinguishes those first-stage computations which are \emph{purely
first-stage} (\bbonep) or contain a \emph{mix} of code from each stage
(\bbonem). This distinction proves useful when defining our splitting algorithm.
%Only mixed code may transition to stage/world \bbtwo.

@@

\crem{Do we actually want to make this ``more efficient'' claim?}

@@

\begin{abstrsyn}

The typing judgment $\typeswor e A$, defined in \ref{fig:statics}, means that
$e$ has type $A$ at world $w$, in the context $\Gamma$. (Variables in the
context are also annotated with the world at which they live.)

All three worlds (\bbonem, \bbonep, and \bbtwo) contain product, sum, function,
and recursive types defined in the usual fashion. For example, given
two terms $\typesone{e_1}A$ and $\typesone{e_2}B$ at world \bbonem, one can form
the pair $\typesone{\texttt{<}e_1,e_2\texttt{>}}{A\times B}$ also at world
\bbonem.

Differing worlds (and hence, differing stages of computation) interact by means
of the $\fut$ and $\curr$ type formers. $\fut A$ is a type in world \bbonem\
which classifies second-stage computations of type $A$. Given a world \bbtwo\
term $e$ of type $A$, $\next{e}$ has type $\fut A$ at world
\bbonem. This essentially encapsulates $e$ as a computation to be evaluated in
the future---first-stage computations can shuffle it around, but not use its
result. The only way to use a $\fut A$ is to wrap it with $\prev$, yielding an
$A$ at world \bbtwo.

%This is why well-typed terms cannot have any information flow from stage
%\bbtwo\ to stage \bbone, and what makes splitting possible.

$\curr A$ is a type in world \bbonem\ which classifies purely-first-stage
computations of type $A$. Given a world \bbonep\ term $e$ of type $A$,
$\pure{e}$ has type $\curr A$ at world \bbonem. ($e$ is guaranteed not to contain
second-stage computations because $\fut$ types are not available in world
\bbonep.)

\crem{Need to explain why we chose letp and caseP as the abstractions.}

\crem{Need to explain why \bbonem\ doesn't have ints.}

\end{abstrsyn}
